---
title: "Final Report"
date: "2023-11-27"
output: 
  html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction


### Movitation

Our motivation stems from the urgent need to comprehensively examine and address homelessness in California. By harnessing the power of data science and leveraging the rich information within HDIS, our project aspires to contribute valuable insights that can inform evidence-based policies and interventions, ultimately paving the way towards a more inclusive and compassionate society.



### Related Work

We are inspired to do this because homelessness has been a long term problem in the United States, which negatively affects cityscape and poses threats to residents’ life and property. Additionally, California has the largest homeless population in the nation. Learning statistical analysis tools from this class, we are motivated to investigate the homelessness situation in California and what are some factors that influence homelessness. 



### Initial Questions

* How is the number of homeless people vary by demographic variables?
* Has the number of homeless people by demographics changed over the year?
* What are some predictors for the number of homeless people?
* How well is our model at predicting the number of homeless people?



## Data Source & Cleaning

We used [data](https://healthdata.gov/State/Hospital-Encounters-for-Homeless-Patients/6ew5-yd9m) from Homelessness Data Integration System (HDIS) which is a publicly available data source from the State of California, Business, Consumer Services, and Housing Agency. The Homelessness Data Integration System (HDIS) is a comprehensive data repository that aggregates and analyzes data from all 44 Continuums of Care (CoC) in California.
We also used another [dataset](https://data.ca.gov/dataset/homelessness-demographics) which included demographic variables including age, race, ethnicity, and gender among from people who are receiving homeless response services in Homelessness Data Integration System.

Firstly, we converted all the variable names to lowercase. Then, we removed the unneeded variable `location_id`, and removed the null value of `experiencing_homelessness`.Finally, we batch export the four cleaned csv files into datasets.The final datasets included the following variables. 

* `CALENDAR_YEAR`: Calendar Year\
* `LOCATION`: Location\
* `ETHNICITY`: Ethnicity\
* `AGE_GROUP_PUBLIC`: Age Group\
* `GENDER`: Gender\
* `RACE`: Race\
* `EXPERIENCING_HOMELESSNESS`: Number of people receiving homelessness services

## Exploratory Analysis
```{r}
library(tidyverse)
library(knitr)
homeless_ip<-read.csv("2019-2020-homeless-ip-and-ed-by-facility.csv")
homeless_data<-homeless_ip %>% 
  filter(HomelessIndicator == "Homeless")
```
#### Description of important variables

`2019-2020-homeless-ip-and-ed-by-facility` contains 17 variables, some of which are listed below:

`Licensed Bed Size`: the hospital's number of licensed beds.                        
`Homeless Indicator`: indicates if the data is for Homeless or Non-Homeless encounters.  
`Demographic Category`: Age, Race, Sex, or (Expected) Payer.       
`Encounters`: Count of inpatient hospitalizations (i.e., discharges) or emergency department visits.         
`Total Hospital Encounters`: Total inpatient hospitalizations or emergency department visits per hospital.     

#### Correlation Matrix of Homeless Data

```{r}
# Correlation matrix
numeric_data <- homeless_data %>% select(Encounters, TotalEncounters, Percent)
correlation_matrix <- cor(numeric_data, method = "pearson")
correlation_table <- kable(correlation_matrix, caption = "Correlation Matrix of Homeless Data")
correlation_table
```
**Encounters and TotalEncounters**: There is a moderate positive correlation of approximately 0.67. This suggests that as the total number of encounters at a facility increases, the number of homeless encounters also tends to increase. This could be expected as larger facilities might see more patients overall, including homeless individuals.

#### Homeless Encounters by Demographic Group (Age, Payer, Race, Sex)

```{r}
homeless_ip<-read.csv("2019-2020-homeless-ip-and-ed-by-facility.csv")

homeless_data<-homeless_ip %>% 
  filter(HomelessIndicator == "Homeless")
combined_plot <- homeless_data %>%
  ggplot(aes(x = DemographicValue, y = Encounters, fill = DemographicValue)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ Demographic, scales = "free_x") + 
  labs(title = "Homeless Encounters by Demographic Group",
       x = "Demographic Value",
       y = "Number of Homeless Encounters") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text.x = element_text(face = "bold")) 
combined_plot
```

This analysis provides a breakdown of homeless encounters by different demographic facets and indicates which groups have higher or lower counts of encounters. It suggests that middle-aged individuals, those covered by Medi-Cal, white individuals, and males have higher numbers of homeless encounters. These findings could point towards specific demographic groups that may require more focused services and interventions. 

#### Homeless Encounters by Bed Size
```{r}
plot2<-homeless_ip %>%
  filter(HomelessIndicator == "Homeless" & EncounterType == "Inpatient Hospitalizations") %>%
  ggplot(aes(x = LicensedBedSize, y = Encounters, fill = LicensedBedSize)) +
  geom_bar(stat = "identity") +
  scale_fill_brewer(palette = "Set3") +
  labs(title = "Homeless Encounters by Bed Size",
       x = "Licensed Bed Size",
       y = "Number of Homeless Encounters") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
plot2
```

Among the variable Bed Size, the category of 400+ Licensed Bed Size shows a high number of encounters of homelessness as compared to other groups, indicating that larger hospitals may have a higher number of such encounters. This could be due to a larger capacity to serve more patients or a higher likelihood of being located in urban areas where homelessness rates may be higher. The second leading category is the 100-199 Licensed Bed Size which was relatively high, because hospitals with 100-199 beds might be specialized in services that are more frequently utilized by the homeless population, such as mental health or substance abuse treatment and these medium-sized hospitals could be strategically located in areas where the homeless population is higher. 

#### Homeless Encounters by Area
```{r}
plot3<-homeless_data %>%
  filter(HomelessIndicator == "Homeless" & EncounterType == "Inpatient Hospitalizations") %>%
  ggplot(aes(x = Urban_Rural, y = Encounters, fill = Urban_Rural)) +
  geom_bar(stat = "identity") +
  scale_fill_brewer(palette = "Set3") +
  labs(title = "Homeless Encounters in Urban vs. Rural Areas",
       x = "Urban/Rural",
       y = "Number of Homeless Encounters") +
  theme_minimal()
plot3
```

This plot help us gain the understanding of the relationship between the number of homelessness encounters and rural/ urban settlement. It is evident that in urban areas there are many people who tend to be homeless due to various reasons such as poverty. Urban hospitals are often larger and have more comprehensive services, which might also contribute to the higher number of encounters. On the other hand, in rural areas less people tend to be homeless since life there is cheap and easily affordable. The more the population in a certain place the higher the chances of homelessness encounters.

#### Homeless Encounters by Ownership Type
```{r}
plot4<-homeless_ip %>%
  filter(HomelessIndicator == "Homeless" & EncounterType == "Inpatient Hospitalizations") %>%
  ggplot(aes(x = Ownership, y = Percent, fill = Ownership)) +
  geom_bar(stat = "identity") +
  scale_fill_brewer(palette = "Set3") +
  labs(title = "Percentage of Homeless Encounters by Ownership Type",
       x = "Ownership",
       y = "Percentage of Homeless Encounters") +
  theme_minimal()
plot4
```

When comparing the percentage of homeless encounters by ownership the non-profit category showed the highest percentage in homeless encounters. Non-Profit hospitals may have more encounters with homeless individuals possibly due to their mission-driven approach, which may include providing care to underserved populations. Investor-owned hospitals, while having a higher percentage of encounters than government hospitals, may still be less than Non-Profit hospitals, potentially due to different operational goals and priorities. Government-owned hospitals having the lowest percentage could be a result of factors such as location, size, the scope of services offered, or specific governmental policies and funding for homeless services.


