---
title: "demo_homeless"
author: "yh wang"
date: "2023-11-04"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import data

This data set gives annual statistics by age group, race, ethnicity, and gender on the number of people who have benefited from homelessness assistance services. 

The Homelessness Data Integration System (HDIS), an extensive data repository that compiles and examines information from each of California's 44 Continuums of Care (CoC), is the source of this information. Every Community of Communities (CoC) collects and disseminates information on the people it serves via a range of programs, such as those aimed at ending homelessness, offering outreach services to the homeless, assisting with permanent housing solutions, and other projects in line with California's Housing First objectives.

```{r}
library(tidyverse)
age_demo = read.csv("./original data/demo/age_demo.csv")
ethnicity_demo = read.csv("./original data/demo/ethnicity_demo.csv")
gender_demo = read.csv("./original data/demo/gender_demo.csv")
race_demo = read.csv("./original data/demo/race_demo.csv")
head(age_demo)
head(ethnicity_demo)
head(gender_demo)
head(race_demo)
```

## Clean data

First, we converted all the variable names to lowercase. 

Then we removed the unneeded variable "location_id", and removed the null value of "experiencing_homelessness".

Finally, we batch export and import the four cleaned csv files into datasets.

```{r}
library(readr)
folder_path <- "./original data/demo/"
csv_files <- list.files(folder_path, pattern = ".csv")

# define a function to process a CSV file
clean_csv <- function(input_file, output_file) {
  data <- read.csv(input_file)
  data_cleaned <- data |> 
    filter(EXPERIENCING_HOMELESSNESS != "*") |>
    select(-LOCATION_ID) |>
    janitor::clean_names() 
   write.csv(data_cleaned, file = output_file, row.names = FALSE)
   #return(data_cleaned)
}

output_folder <- "./processed data/demo/"

# process CSV files using the lapply function
cleaned_data_list <- lapply(csv_files, function(file) {
  input_file <- paste0(folder_path, file)
  output_file <- paste0(output_folder, "cleaned_", file)
  clean_csv(input_file, output_file)
})

# import clean data
age_clean = read.csv("./processed data/demo/cleaned_age_demo.csv")
ethnicity_clean = read.csv("./processed data/demo/cleaned_ethnicity_demo.csv")
gender_clean = read.csv("./processed data/demo/cleaned_gender_demo.csv")
race_clean = read.csv("./processed data/demo/cleaned_race_demo.csv")
```

## Data visualization

The data visualization serves as a foundation for a more in-depth investigation by providing us with a preliminary understanding of the relationship between the variables (age, gender, ethnicity, and race) and the number of homeless people and how that relationship has evolved over time.

```{r}
# line plot for "Total Homelessness by Year and Age Group"
library(plotly)
age_plot <- age_clean |>
  select(-location) |>
  group_by(calendar_year,age_group_public) |>
  summarise(total_homelessness = sum(experiencing_homelessness))

age_plot |> 
  mutate(text_label = str_c("Year: ", calendar_year, "\nAge Group: ", age_group_public)) |> 
  plot_ly(x = ~calendar_year, y = ~total_homelessness, type = "scatter", mode = "line", color = ~age_group_public, colors = "viridis", text = ~text_label, alpha = 0.8) |>
    layout(title = "Total Homelessness by Year and Age Group")
```

```{r}
# bar chart for "Total Homelessness by Year and ethnicity"
ethnicity_plot <- ethnicity_clean |>
  select(-location) |>
  group_by(calendar_year,ethnicity) |>
  summarise(total_homelessness = sum(experiencing_homelessness))

ethnicity_plot |> 
  mutate(text_label = str_c("Year: ", calendar_year, "\nEthnicity: ", ethnicity)) |>
  plot_ly(x = ~calendar_year, y = ~total_homelessness, type = "bar", color = ~ethnicity, colors = "viridis", text = ~text_label, alpha = 0.8) |>
  layout(
    title = "Total Homelessness by Year and ethnicity")
```

```{r}
# bar chart for "Total Homelessness by Year and gender"
gender_plot <- gender_clean |>
  select(-location) |>
  group_by(calendar_year,gender) |>
  summarise(total_homelessness = sum(experiencing_homelessness))

gender_plot  |> 
  mutate(text_label = str_c("Year: ", calendar_year, "\nGender: ", gender)) |>
  plot_ly(x = ~calendar_year, y = ~total_homelessness, type = "bar", color = ~gender, colors = "viridis", text = ~text_label, alpha = 0.8) |>
  layout(
    title = "Total Homelessness by Year and gender")
```

```{r}
# line plot for "Total Homelessness by Year and gender"
race_plot <- race_clean |>
  select(-location) |>
  group_by(calendar_year,race) |>
  summarise(total_homelessness = sum(experiencing_homelessness))

race_plot  |> 
  mutate(text_label = str_c("Year: ", calendar_year, "\nRace: ", race)) |>
  plot_ly(x = ~calendar_year, y = ~total_homelessness,  type = "scatter", mode = "line", color = ~race, colors = "viridis", text = ~text_label, alpha = 0.8) |>
  layout(
    title = "Total Homelessness by Year and gender" ,
    legend = list(
      x = 1,
      y = 0.5, 
      traceorder = "normal",
      bgcolor = "white", 
      bordercolor = "white",
      borderwidth = 0.5
    ))

# pie chart for "Percentage of Total Homeless by Race"
race_clean |>
  group_by(race) |>
  summarise(total_homelessness = sum(experiencing_homelessness)) |>
  mutate(percentage = total_homelessness / sum(total_homelessness)) |>
  plot_ly(labels = ~race, values = ~percentage, type = "pie", hole = 0.4) |>
   layout(title = "Percentage of Total Homeless by Race")
```