---
title: "Final Project Shiny Dashboard"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    source_code: embed
runtime: shiny
---

```{r setup, include=FALSE}
library(flexdashboard)
library(tidyverse)
library(shiny)
library(plotly)

homeless_data<-read.csv("data/2019-2020-homeless-ip-and-ed-by-facility.csv")
age_clean = read.csv("data/cleaned_age_demo.csv")
ethnicity_clean = read.csv("data/cleaned_ethnicity_demo.csv")
gender_clean = read.csv("data/cleaned_gender_demo.csv")
race_clean = read.csv("data/cleaned_race_demo.csv")
```

Homeless People Receiving Hospital Healthcare Services in Hospitals
===================================== 

Column {.sidebar}
-----------------------------------------------------------------------

```{r}
DemoValue = homeless_data |> distinct(Demographic) |> pull()

radioButtons(
  inputId = "DemoValue", 
  label = h3("Select a Demographic Category"),
  choices = DemoValue,
  selected = "Age")
```

Column {data-height=650}
-----------------------------------------------------------------------

### Description

While this dataset is rich in hospital encounter information, it seems to differ from the ‘demo’ datasets described earlier. 

\ \par

The ‘demo’ datasets focus on broader statistics of people receiving homeless response services across various dimensions such as age, race, ethnicity, and gender, 

\ \par

and are sourced from the Homelessness Data Integration System (HDIS) which compiles data from California’s Continuums of Care. 

\ \par

In contrast, the provided dataset is more focused on hospital encounters of homeless individuals, offering a detailed view of healthcare utilization among this population.

Column {data-height=350}
-----------------------------------------------------------------------

### Bar Plot

```{r}
homeless_data_bar =
  homeless_data |>
  group_by(Demographic, DemographicValue) |>
  summarise(total_homelessness = sum(Encounters))

renderPlotly({
  homeless_data_bar |> 
    filter(Demographic == input$DemoValue) |>
    plot_ly(x = ~DemographicValue, y = ~total_homelessness, type = 'bar') |>
    layout(
      title = "Total Homelessness Receiving Hospitals' Services")
})
```

Homeless People by Age
===================================== 

Column {.sidebar}
-----------------------------------------------------------------------

```{r}
year = age_clean %>% distinct(calendar_year) %>% pull()

# selectInput widget
selectInput(
  inputId = "year", 
  label = h3("Select Year"),
  choices = year,
  selected = "2017")
```

Column {data-height=650}
-----------------------------------------------------------------------

### Line Plot

```{r}
age_line =
  age_clean |>
  select(-location) |>
  group_by(calendar_year, age_group_public) |>
  summarise(total_homelessness = sum(experiencing_homelessness))

renderPlotly({
age_line |> 
  mutate(text_label = str_c("Year: ", calendar_year, "\nAge Group: ", age_group_public)) |> 
  plot_ly(x = ~calendar_year, y = ~total_homelessness, type = "scatter", mode = "line", color = ~age_group_public, colors = "viridis", text = ~text_label, alpha = 0.8) |>
    layout(title = "Total Homelessness by Age")

})
```

Column {data-height=350}
-----------------------------------------------------------------------

### Bar Plot

```{r}
age_bar = 
  age_clean |>
  group_by(calendar_year, age_group_public) |>
  summarise(total_homelessness = sum(experiencing_homelessness))

renderPlotly({
  age_bar |> 
    filter(calendar_year == input$year) |>
    plot_ly(x = ~age_group_public, y = ~total_homelessness, type = "bar") |>
    layout(
      title = "Total Homelessness by Age across Years")
})
```

Homeless People by Race
===================================== 

Column {.sidebar}
-----------------------------------------------------------------------

```{r}
year2 = race_clean %>% distinct(calendar_year) %>% pull()

# selectInput widget
selectInput(
  inputId = "year2", 
  label = h3("Select Race"),
  choices = year2,
  selected = "2017")

```

Row{data-height=650}
-----------------------------------------------------------------------

### Line Plot

```{r}
race_line =
  race_clean |>
  select(-location) |>
  group_by(calendar_year, race) |>
  summarise(total_homelessness = sum(experiencing_homelessness))

renderPlotly({
race_line |> 
  mutate(text_label = str_c("Year: ", calendar_year, "\nRace Group: ", race)) |> 
  plot_ly(x = ~calendar_year, y = ~total_homelessness, type = "scatter", mode = "line", color = ~race, colors = "viridis", text = ~text_label, alpha = 0.8) |>
    layout(title = "Total Homelessness by Race")

})
```

Row{data-height=350}
-----------------------------------------------------------------------

### Bar Plot

```{r}
race_box = 
  race_clean |>
  group_by(calendar_year,race) |>
  summarise(total_homelessness = sum(experiencing_homelessness))

renderPlotly({
  race_box |> 
    filter(calendar_year == input$year2) |>
    plot_ly(x = ~race, y = ~total_homelessness, type = "bar") |>
    layout(
      title = "Total Homelessness by Race across Years")
})
```

Homeless People by Ethnicity
===================================== 

Column {.sidebar}
-----------------------------------------------------------------------

```{r}
year3 = ethnicity_clean %>% distinct(calendar_year) %>% pull()

# selectInput widget
selectInput(
  inputId = "year3", 
  label = h3("Select Ethnicity"),
  choices = year3,
  selected = "2017")

```

Row{data-height=650}
-----------------------------------------------------------------------

### Line Plot

```{r}
ethnicity_line =
  ethnicity_clean |>
  select(-location) |>
  group_by(calendar_year, ethnicity) |>
  summarise(total_homelessness = sum(experiencing_homelessness))

renderPlotly({
ethnicity_line |> 
  mutate(text_label = str_c("Year: ", calendar_year, "\nEthnicity Group: ", ethnicity)) |> 
  plot_ly(x = ~calendar_year, y = ~total_homelessness, type = "scatter", mode = "line", color = ~ethnicity, colors = "viridis", text = ~text_label, alpha = 0.8) |>
    layout(title = "Total Homelessness by Ethnicity")

})
```

Row{data-height=350}
-----------------------------------------------------------------------

### Bar Plot

```{r}
ethnicity_box = 
  ethnicity_clean |>
  group_by(calendar_year,ethnicity) |>
  summarise(total_homelessness = sum(experiencing_homelessness))

renderPlotly({
  ethnicity_box |> 
    filter(calendar_year == input$year3) |>
    plot_ly(x = ~ethnicity, y = ~total_homelessness, type = "bar") |>
    layout(
      title = "Total Homelessness by Ethnicity across Years")
})
```

Homeless People by Gender
===================================== 

Column {.sidebar}
-----------------------------------------------------------------------

```{r}
year4 = gender_clean %>% distinct(calendar_year) %>% pull()

# selectInput widget
selectInput(
  inputId = "year4", 
  label = h3("Select Gender"),
  choices = year4,
  selected = "2017")

```

Row{data-height=650}
-----------------------------------------------------------------------

### Line Plot

```{r}
gender_line =
  gender_clean |>
  select(-location) |>
  group_by(calendar_year, gender) |>
  summarise(total_homelessness = sum(experiencing_homelessness))

renderPlotly({
gender_line |> 
  mutate(text_label = str_c("Year: ", calendar_year, "\nGender Group: ", gender)) |> 
  plot_ly(x = ~calendar_year, y = ~total_homelessness, type = "scatter", mode = "line", color = ~gender, colors = "viridis", text = ~text_label, alpha = 0.8) |>
    layout(title = "Total Homelessness by Gender")

})
```

Row{data-height=350}
-----------------------------------------------------------------------

### Bar Plot

```{r}
gender_box = 
  gender_clean |>
  group_by(calendar_year,gender) |>
  summarise(total_homelessness = sum(experiencing_homelessness))

renderPlotly({
  gender_box |> 
    filter(calendar_year == input$year4) |>
    plot_ly(x = ~gender, y = ~total_homelessness, type = "bar") |>
    layout(
      title = "Total Homelessness by Gender across Years")
})
```
