---
title: "reg models"
author: "Yiying Wu"
date: "2023-11-26"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(corrplot)
```

## Import Data & Recoding
First import data and create dummy variables for categorical variables.
```{r, message=FALSE}
dat <- read_csv("2019-2020-homeless-ip-and-ed-by-facility.csv")

### EncounterType
dat$ed_visit <- ifelse(dat$EncounterType=="ED Visits",1,0)

### Ownership
dat$government <- ifelse(dat$Ownership=="Government",1,0)
dat$investor <- ifelse(dat$Ownership=="Investor",1,0)

### Urban_Rural
dat$rural <- ifelse(dat$Urban_Rural=="Rural/Frontier",1,0)

### LicensedBedSize
## the first level will be treated as ref
dat$LicensedBedSize <- as.factor(dat$LicensedBedSize)
dat$ordered_LicensedBedSize <- ordered(dat$LicensedBedSize, 
                             levels = c("1-99","100-199","200-299","300-399","400+"))
```
The variables and corresponding descriptions are as follows.

| Variable | Description | Value |
|-------|-----------|-----------|
| ed_visit | Dummy variable for EncounterType  | 1=ED Visits, 0=Inpatient Hospitalizations   |
| government | Dummy variable for Ownership | 1=Government, 0=Otherwise |
| investor | Dummy variable for Ownership | 1=Investor, 0=Otherwise |
| rural | Dummy variable for Urban_Rural | 1=Rural/Frontier, 0=Urban |

## Univariate Analysis
We conducted univariate analyses to identify individual factors associated with outcomes `Encounters`. 

```{r}
# Function to round all numeric columns in a dataframe
round_df <- function(df, digits) {
  df %>% mutate(across(where(is.numeric), ~ round(., digits)))
}

# Create tidied summaries for each model and round them
### EncounterType
tidy_ed_visit <- round_df(broom::tidy(lm(Encounters ~ ed_visit, data = dat)), 4) %>%
  mutate(Model = "EncounterType")

### Ownership
tidy_Ownership <- round_df(broom::tidy(lm(Encounters ~ government + investor, data = dat)), 4) %>%
  mutate(Model = "Ownership")

### Urban_Rural
tidy_Urban_Rural <- round_df(broom::tidy(lm(Encounters ~ rural, data = dat)), 4) %>%
  mutate(Model = "Urban_Rural")

### LicensedBedSize
tidy_LicensedBedSize <- round_df(broom::tidy(lm(Encounters ~ ordered_LicensedBedSize, data = dat)), 4) %>%
  mutate(Model = "LicensedBedSize")

# Combine all tidied summaries into one table
combined_table <- bind_rows(tidy_ed_visit, tidy_Ownership, tidy_Urban_Rural, tidy_LicensedBedSize)

# Output the table
knitr::kable(combined_table)
```
**Description**

* **EncounterType Model**: The increase in the number of encounters for each additional emergency department visit. It is highly significant (p < 0.0001), suggesting a strong relationship between emergency department visits and the number of encounters.
* **Ownership Model**: The differences in the number of encounters for government-owned and investor-owned facilities compared to the Non-Profit facilities. Both coefficients are significant (p < 0.0001), indicating that ownership type is an important predictor of encounters. 
* **Urban_Rural Model**: The difference in the number of encounters between Rural/Frontier and Urban facilities. The negative coefficient suggests that Rural/Frontier facilities have fewer encounters than urban ones, and this is highly significant (p < 0.0001).
* **LicensedBedSize Model**: Coefficients (L, Q, C, ^4) represent the linear, quadratic, cubic, and quartic terms in a polynomial regression model for ordered bed sizes. The 'L' term has a highly significant positive coefficient, indicating an initial increase in encounters with bed size. However, the lack of significance in the quadratic ('Q') and cubic ('C') terms, and the non-significant quartic term ('^4'), suggests that the relationship might not be strongly non-linear or that there is insufficient data to detect higher-order non-linearities.


## Correlation Matrix
Since `ordered_LicensedBedSize` is ordinal data, we use Spearman correlation coefficients when constructing correlation matrix.
```{r,message=FALSE}
dat1 <- dat|>
  mutate(ordered_LicensedBedSize=as.numeric(ordered_LicensedBedSize))|>
  select(ed_visit,government,investor,rural,ordered_LicensedBedSize,Encounters)

correlation_matrix <- cor(dat1, method = "spearman", use = "complete.obs")

# Visualize
corrplot(correlation_matrix, method = "color", addCoef.col = "black", tl.col = "black", tl.srt = 45, insig = "blank" , number.cex = 0.7, diag = FALSE)
```

**Description**

* the variable ordered_LicensedBedSize seems to have a moderately positive correlation with Encounters (0.34), suggesting that facilities with more licensed beds tend to have more encounters.
* rural has a strong negative correlation with ordered_LicensedBedSize (-0.45), implying that rural facilities are likely to have fewer licensed beds.
* Correlation between all variables are below 70%, indicating that it is less likely that collinearity will pose a problem for the regression model.
