---
title: "reg models"
author: "Yiying Wu"
date: "2023-11-26"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(corrplot)
```

## Import Data & Recoding
First import data and create dummy variables for categorical variables.
```{r, message=FALSE}
dat <- read_csv("2019-2020-homeless-ip-and-ed-by-facility.csv")

### EncounterType
dat$ed_visit <- ifelse(dat$EncounterType=="ED Visits",1,0)

### Ownership
dat$government <- ifelse(dat$Ownership=="Government",1,0)
dat$investor <- ifelse(dat$Ownership=="Investor",1,0)

### Urban_Rural
dat$rural <- ifelse(dat$Urban_Rural=="Rural/Frontier",1,0)

### LicensedBedSize
## the first level will be treated as ref
dat$LicensedBedSize <- as.factor(dat$LicensedBedSize)
dat$ordered_LicensedBedSize <- ordered(dat$LicensedBedSize, 
                             levels = c("1-99","100-199","200-299","300-399","400+"))
```
The variables and corresponding descriptions are as follows.

| Variable | Description | Value |
|-------|-----------|-----------|
| ed_visit | Dummy variable for EncounterType  | 1=ED Visits, 0=Inpatient Hospitalizations   |
| government | Dummy variable for Ownership | 1=Government, 0=Otherwise |
| investor | Dummy variable for Ownership | 1=Investor, 0=Otherwise |
| rural | Dummy variable for Urban_Rural | 1=Rural/Frontier, 0=Otherwise |
| ordered_LicensedBedSize.L | Dummy variable for LicensedBedSize | 1=100-199, 0=Otherwise |
| ordered_LicensedBedSize.Q | Dummy variable for LicensedBedSize | 1=200-299, 0=Otherwise |
| ordered_LicensedBedSize.C | Dummy variable for LicensedBedSize | 1=300-399, 0=Otherwise |
| ordered_LicensedBedSize^4 | Dummy variable for LicensedBedSize | 1=400+, 0=Otherwise |

## Univariate Analysis
We conducted univariate analyses to identify individual factors associated with outcomes `Encounters`. 

```{r}
# Function to round all numeric columns in a dataframe
round_df <- function(df, digits) {
  df %>% mutate(across(where(is.numeric), ~ round(., digits)))
}

# Create tidied summaries for each model and round them
### EncounterType
tidy_ed_visit <- round_df(broom::tidy(lm(Encounters ~ ed_visit, data = dat)), 4) %>%
  mutate(Model = "EncounterType")

### Ownership
tidy_Ownership <- round_df(broom::tidy(lm(Encounters ~ government + investor, data = dat)), 4) %>%
  mutate(Model = "Ownership")

### Urban_Rural
tidy_Urban_Rural <- round_df(broom::tidy(lm(Encounters ~ rural, data = dat)), 4) %>%
  mutate(Model = "Urban_Rural")

### LicensedBedSize
tidy_LicensedBedSize <- round_df(broom::tidy(lm(Encounters ~ ordered_LicensedBedSize, data = dat)), 4) %>%
  mutate(Model = "LicensedBedSize")

# Combine all tidied summaries into one table
combined_table <- bind_rows(tidy_ed_visit, tidy_Ownership, tidy_Urban_Rural, tidy_LicensedBedSize)

# Output the table
knitr::kable(combined_table)
```


## Correlation Matrix
Since `ordered_LicensedBedSize` is ordinal data, we use Spearman correlation coefficients when constructing Correlation Matrix.
```{r,message=FALSE}
dat1 <- dat|>
  mutate(ordered_LicensedBedSize=as.numeric(ordered_LicensedBedSize))|>
  select(ed_visit,government,investor,rural,ordered_LicensedBedSize,Encounters)

correlation_matrix <- cor(dat1, method = "spearman", use = "complete.obs")

# Visualize
corrplot(correlation_matrix, method = "color", addCoef.col = "black", tl.col = "black", tl.srt = 45, insig = "blank" , number.cex = 0.7, diag = FALSE)
```

